---
pagetitle: 'Exercise: Structure Populations - Sensitivity II'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# Structure Populations - Sensitivity Analysis II {.tabset}

In the previous exercise, you expolored numerically how population growth and the dominant eigenvalue $\lambda$ of a population projection matrix $\sf L$ changes in response to changes in the different matrix entries, i.e., the different demographic parameters. But this way of anaylyzing the effect of changes in the different demograhic parameters on $\lambda$ has the disadvantage that the result depends on the actual magnitude by which the matrix entries are increased (e.g., $0.1$ or $40\%$).

An alternative method is *sensitivty analysis*. In this method, we compute the derivative of the dominant eigenvalue $\lambda$ with respect to the different matrix entries. In other words, we calculate the slope of $\lambda$ as a function of each of the different matrix entries. This is what we will do in the this exercise.

### Sensistivity analysis

Recall from the lecture that the sensitivity of the dominant eigenvalue with respect to an arbitrary matrix entry $a_{ij}$ is given by

$$
s_{ij} = \frac{\partial\lambda}{\partial a_{ij}}=\frac{v_j u_i}{<\mathbf v, \mathbf u>},
$$

where $\mathbf u=(u_1, \ldots, u_n)$ and $\mathbf v=(v_1, \ldots, v_n)$ are the right and left eigenvector of an $n$-dimensional population projection matrix $\mathsf A$ and give the stable age distribution and the vector of reproductive values, respectively. Furthermore, $<\mathbf v, \mathbf u>$ is the *inner product* of two vectors defined as

$$
<\mathbf v, \mathbf u> = v_1u_1+\ldots+v_nu_n.
$$
To compute the whole sensitivity matrix in one step we can make use of the *outer product* of two vectors defined as

$$
\mathbf v^{\mathrm T} \mathbf u = 
\begin{pmatrix}
v_1\\
v_2\\
\vdots\\
v_n
\end{pmatrix}
\begin{pmatrix}
u_1 & u_2 \ldots & u_1
\end{pmatrix}=
\begin{pmatrix}
v_1u_1 & \ldots & v_1u_n\\
v_2u_1 & \ldots & v_2u_n\\
\vdots & \ddots & \vdots\\
v_nu_1 & \ldots & v_nu_n
\end{pmatrix}
$$

The sensitity matrix $\mathsf S$ can then be computed as

$$
\mathsf S=\frac{\mathbf v^{\mathrm T} \mathbf u}{<\mathbf v, \mathbf u>}.
$$

Thus, in order to calculate the sensitivity matrix $\sf S$, we first have to calculate the right and left eigenvector of the population projection matrix $\sf L$ corresponding to the dominant eigenvalue $\lambda$. The right eigenvector gives the *stable age-distribution*, which we have calculated already in the first exercise. The following code simply repeats this calculation.

```{r}
L <- matrix(
  c(0,1 ,5, 0.25, 0, 0, 0, 0.5, 0),	 # the matrix entries
  nrow = 3, # number of rows
  ncol = 3, # number of columns
  byrow = T) # fill matrix by rows

N.0 <- c(2,2,1) # vector of initial densities
length <- 40 # lenght of the time series
```

```{r eval=FALSE}
time.series <- matrix(N.0) # stores the initial population vector N.0 in the matrix "time.series"
for (i in 1:length){# executes the multiplication of L.N(t) as often as specified by the parameter "length"
  time.series <- cbind(time.series, L %*% time.series[ ,i])
}

matplot(0:length, t(time.series), type="l", lty=1:3, col = 1:3, ylab = "abundance", xlab = "time", main="original matrix") # plots result
legend("topright", legend = c("age 1", "age 2", "age 3"), lty = 1:3, col = 1:3, bty = "n") # defines legend

lambda <- time.series[1, length+1]/time.series[1, length] # ratio of population size at the end of the time series for age-class 1 gives lambda
lambda # print dominant eigenvalue

time.series[ , length+1] # vector of population sizes at the end of the time series gives the right eigenvector. In order to express the right eigenvector in terms of proportions we divided each entry by the total population size.

N.tot <- sum(time.series[ , length+1]) # sum of the densities contained in the population vector at the end of the time series.

right.EV <- time.series[ , length+1]/N.tot # right eigenvector standardized such that its entries sum up to one. This vector represents proportions in the stable age distribution.
right.EV # print stable age distribution
```

In order to obtain the left eigenvector, that is, the vector of reproductive values, we follow exacty the same procedure but using the transpose $\sf L^{\mathrm{T}}$ of the population projection matrix.

```{r eval=FALSE}
L.transpose <- t(L) # transpose of the population projection matrix L
time.series.transpose <- matrix(N.0)

for (i in 1:length){# executes the multiplication of L.N.transpose(t) as often as specified by the parameter "length"
  time.series.transpose <- cbind(time.series.transpose, L.transpose %*% time.series.transpose[ ,i])
}

matplot(0:length, t(time.series.transpose), type="l", lty=1:3, col = 1:3, ylab = "something that is NOT density", xlab = "time", main="transposed matrix") # plots result, note that this graph has NO biological interpretation, it is only a tool to obtain the left eigenvalue
legend("topright", legend = c("age 1", "age 2", "age 3"), lty = 1:3, col = 1:3, bty = "n") # defines legend

left.EV <- time.series.transpose[ , length+1]/time.series.transpose[1, length+1] # left eigenvector containing reproductive values with the reproductive value for the first age-class standardized to 1.

left.EV # print left eigenvector
```

We now have all ingredients to compute the sensitivity matrix $\sf S$.

```{r eval=FALSE}
inner.prod <-left.EV %*% right.EV # this corresponds to <v, u>
inner.prod # prints the inner product
outer.prod <- outer(left.EV, right.EV) # this corresponds to v^T*u
outer.prod # prints the outer product
sensitivities <- outer.prod/inner.prod[1, 1] # note that by writing inner.prod[1, 1] we extract the value from a "one-dimensional matrix"
sensitivities # prints the sensitivity matrix
```

Each entry of this matrix contains the sensitivity of $\lambda$ corresponding to the matrix entry at that particular position.

  1) Note that this matrix has only non-zero entries while our population projection matrix $\sf L$ has five entries that are equal to zero. What is the meaning of the sensitivity of matrix entry that is zero? For example, in $\sf L$ fecundity of one year old individuals equals zero, $f_1=0$. But the derivative of $\lambda$ with respect to $f_1$ equals $0.9023$. What does this mean?
  
  2) What conclusion do you draw from the sensitivity matrix $\sf S$ from a conservation perspective?
  
  3) Based on $\sf S$, estimate by how much $\lambda$ increases if $f_3$ is increased by $0.5$, $1$ and $2$. Similarly,estimate by how much $\lambda$ increases if $s_1$ is increased by $0.1$ or $0.2$.
  
  4) How does your prediction based on the sensitivity matrix $\sf S$ compare to the previous exercise when you increased all matrix entries equally by a value of $0.1$?

### Elasticity analysis

Sensitivies give the absolute change in $\lambda$ given an absolute change in the demograhic parameters $a_{ij}$. Thus, sensitivies do not take into account that survival probabilities and fecundities can be on very different scales. To correct for this, we can compute the *elasticity matrix* $\sf E$, defined as

$$
\mathsf E=\frac{a_{ij}}{\lambda}\mathsf S=\frac{a_{ij}}{\lambda}\frac{\mathbf v^{\mathrm T} \mathbf u}{<\mathbf v, \mathbf u>}.
$$

This gives the *proportional change* in $\lambda$ given a proportional change in the demographic parameter $a_{ij}$. In other words, it gives by how many percentages $\lambda$ changes given a change in percentage of the matrix entry $a_{ij}$. We can compute the elasity matrix as follows:

```{r eval=FALSE}
elasticities <- (1/lambda)*L*sensitivities # matrix of elasticities. Here, L*sensitivities is the "element-wise product matrix"
elasticities # print elasticities
```

Two observations are noteworthy about the elasiticity matrix $\sf E$. First, for matrix entries $a_{ij}=0$ the corresponding elastity value $e_{ij}$ equals zero as well. This is because we multiply each sensitivity value with the correspoding value $a_{ij}$ when computing the elasticities. Second, the entries of the elasticity matrix $\sf E$ sum up to one. This indicates that its entries indeed give the proportional effect of each matrix entry $a_{ij}$ on $\lambda$.

  1) How does your prediction based on the elasticity matrix $\sf E$ compare to the previous exercise when you increased all matrix entries equally by $40 \%$?